---
title: "Differentiating Through a Cone Program with diffcp"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Differentiating Through a Cone Program with diffcp}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`diffcp` solves a convex cone program

$$
\begin{aligned}
\text{minimise}   &\quad c^\top x \\
\text{subject to} &\quad Ax + s = b, \quad s \in K
\end{aligned}
$$

and returns both the optimum $(x, y, s)$ (with $y$ the dual variable
of the cone constraint) **and** an abstract linear map
representing the derivative of $(x, y, s)$ with respect to the data
$(A, b, c)$. The derivative is exposed as two callables:

* `D(dA, db, dc)` — the forward derivative; returns `(dx, dy, ds)`.
* `DT(dx, dy, ds)` — the adjoint; returns `(dA, db, dc)`.

This vignette walks through three small examples that exercise the
common cones and demonstrate both the forward solve and the
derivative.

```{r setup}
library(diffcp)
library(Matrix)
```

## A linear program

The simplest case: a 1-equality / non-negative orthant LP.

$$
\min_{x \in \mathbb{R}^3} \; c^\top x
\quad \text{s.t.} \quad
\mathbf{1}^\top x = 1, \; x \ge 0.
$$

In SCS-style standard form $Ax + s = b$, $s \in K = \{0\} \times
\mathbb{R}^3_+$:

```{r lp-data}
A <- sparseMatrix(i = c(1, 2, 1, 3, 1, 4),
                  j = c(1, 1, 2, 2, 3, 3),
                  x = c(1, -1, 1, -1, 1, -1),
                  dims = c(4, 3))
b <- c(1, 0, 0, 0)
c <- c(1, 2, 3)
cone_dict <- list(z = 1L, l = 3L)
```

Solve:

```{r lp-solve}
res <- solve_and_derivative(A, b, c, cone_dict, mode = "lsqr",
                            tol_gap_abs = 1e-10,
                            tol_gap_rel = 1e-10,
                            tol_feas    = 1e-10)
res$x
```

As expected the optimum is at $x = (1, 0, 0)$ (the cheapest non-zero
coordinate). Now apply the derivative at a small perturbation in $c$:

```{r lp-D}
dA <- sparseMatrix(i = integer(0), j = integer(0), x = numeric(0),
                   dims = c(4, 3))
db <- numeric(4)
dc <- c(0.001, 0, 0)
d <- res$D(dA, db, dc)
d$dx
```

This should agree with finite-differencing the perturbed solve:

```{r lp-fd}
res_pert <- solve_only(A, b, c + dc, cone_dict,
                       tol_gap_abs = 1e-10,
                       tol_gap_rel = 1e-10,
                       tol_feas    = 1e-10)
max(abs(d$dx - (res_pert$x - res$x)))
```

## A second-order cone program

Minimum-distance-to-the-simplex inside a Lorentz cone:

$$
\min_{t, x} \; t \quad \text{s.t.} \quad \|x\|_2 \le t, \; \mathbf{1}^\top x = 1.
$$

Variables: $(t, x_1, x_2, x_3)$. Standard form $Ax + s = b$ with one
zero cone (the equality) and one SOC of size 4 (containing the
$(t, x)$ block):

```{r soc-data}
A <- sparseMatrix(
  i = c(2, 1, 3, 1, 4, 1, 5),
  j = c(1, 2, 2, 3, 3, 4, 4),
  x = c(-1, 1, -1, 1, -1, 1, -1),
  dims = c(5, 4))
b <- c(1, 0, 0, 0, 0)
c <- c(1, 0, 0, 0)
cone_dict <- list(z = 1L, q = 4L)

res <- solve_and_derivative(A, b, c, cone_dict, mode = "dense",
                            tol_gap_abs = 1e-10,
                            tol_gap_rel = 1e-10,
                            tol_feas    = 1e-10)
res$x
```

The optimum has $x = (1/3, 1/3, 1/3)$ and $t = \sqrt{3}/3 \approx
0.577$, achieving the SOC equality constraint $\|x\|_2 = t$.

## A semidefinite program

Minimum-eigenvalue of a 5×5 PSD matrix subject to unit trace. Since
PSD vectorisation in SCS / `diffcp` follows lower-triangular
column-major order with `sqrt(2)` scaling on the off-diagonals, we
build the trace-coefficient row by hand:

```{r sdp-data}
DIM <- 5L
n   <- DIM * (DIM + 1L) %/% 2L          # 15

trace_row <- numeric(n)
pos <- 1L
for (col in seq_len(DIM)) {
  trace_row[pos] <- 1.0
  pos <- pos + (DIM - col + 1L)         # advance to next diagonal entry
}

A <- rbind(
  Matrix(matrix(trace_row, nrow = 1L), sparse = TRUE),
  -Diagonal(n))
A <- as(A, "CsparseMatrix")
b <- c(1.0, numeric(n))

C <- matrix(0, DIM, DIM)
C[1, 1] <- 1; C[DIM, DIM] <- -1
c_vec <- vec_symm(C)

cone_dict <- list(z = 1L, s = DIM)

res <- solve_only(A, b, c_vec, cone_dict,
                  tol_gap_abs = 1e-10,
                  tol_gap_rel = 1e-10,
                  tol_feas    = 1e-10)
X <- unvec_symm(res$x, DIM)
sum(diag(X))
range(eigen(X, symmetric = TRUE, only.values = TRUE)$values)
```

The optimal $X$ has trace 1 and concentrates on the eigenvector of
$C$ with the most-negative eigenvalue, so $X \approx e_5 e_5^\top$.

## Choosing the derivative mode

`solve_and_derivative` accepts two modes:

* `mode = "lsqr"` (default): matrix-free conjugate-gradient style
  iterative solve against the `M` operator. Memory-efficient and
  appropriate for large sparse problems.
* `mode = "dense"`: build `M` as a dense matrix and factor with
  Eigen's LDLT. Lower per-call cost when the problem is small or the
  derivative is being applied many times to different perturbations.

Both modes produce the same numeric answer to within solver
tolerance; the dense path matches Python's
`(M^\top M).\text{ldlt}().\text{solve}(M^\top \cdot \text{rhs})` bit
for bit, and the LSQR path agrees with the dense path to roughly
`1e-6` (LSQR's default `atol = btol = 1e-8`).

## Citation

If you use `diffcp` in academic work, please cite:

> Agrawal, A.; Barratt, S.; Boyd, S.; Busseti, E.; Moursi, W.
> "Differentiating through a cone program."
> *Journal of Applied and Numerical Optimization*, **1**(2), 107–115, 2019.
